\documentclass[12pt]{clapps}  % Praxia/Kontablo house style (clean-room)

% --- PACKAGES ---
% geometry, lmodern, microtype, xcolor and hyperref are provided by clapps.cls.
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{float}
\usepackage{url}
% Breakable monospace path macro: long inline file paths wrap at separators
% (using the aggressive \UrlBreaks set defined in clapps.cls) instead of
% overflowing the right margin. Underscores are written literally inside it.
\DeclareUrlCommand\pathtt{\urlstyle{tt}}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows.meta, positioning, fit, backgrounds}

% --- PREAMBLE ---
\title{Kontablo: A Graph-Based Universal Accounting Ontology for the M2M Agentic Economy}
\author{Christian Luciani}
\date{\small{Version 1.9.2 --- June 2026}}

\begin{document}

\maketitle

\clappsepigraph{Confusion of tongues did not stop the building; it only made
every later floor cost more to align.}{On the Accounting Babel}

% --- SECTIONS ---
\input{sections/abstract}

\compactcontents

\input{sections/summary_instructions}

\input{sections/summary_tables}

\input{sections/introduction}

\input{sections/precedents}

\input{sections/impact}

\input{sections/ontology}

\input{sections/mathematical_foundations}

\input{sections/related_work}

\input{sections/deterministic_logic}

\input{sections/harness_architecture}

\input{sections/agentic_economy}

\input{sections/governance}

\input{sections/evaluation}

\input{sections/remarks}

\input{sections/labor_market}

\input{sections/conclusion}

\input{sections/acknowledgements}

\input{sections/appendix}

\input{sections/glossary}

% --- BIBLIOGRAPHY ---
\section*{Selected Bibliography}
\begin{itemize}
    \item IFRS Foundation. (2024). \textit{IFRS Taxonomy 2024}.
    \item IFRS Foundation. (2025). \textit{Who Uses IFRS Accounting Standards?} \url{https://www.ifrs.org/use-around-the-world/use-of-ifrs-standards-by-jurisdiction/} [Profiles for 169 jurisdictions.]
    \item International Accounting Standards Board. \textit{IAS 21 --- The Effects of Changes in Foreign Exchange Rates}. IFRS Foundation. [Foreign-currency transactions recorded at the spot rate at the transaction date; average-rate practical expedient permitted where rates do not fluctuate significantly.]
    \item European Securities and Markets Authority. (2023). \textit{European Single Electronic Format (ESEF) Reporting Manual}. ESMA. [Mandated inline-XBRL (iXBRL) disclosure format for EU-regulated issuers, tagged against the IFRS Taxonomy.]
    \item SAP SE. (2024). \textit{SAP S/4HANA Finance: the Universal Journal (ACDOCA)}. [Single unified line-item table consolidating GL, controlling, and reporting for multi-GAAP, multi-currency postings within one instance.]
    \item Oracle Corporation. (2024). \textit{Oracle ERP: Primary, Secondary, and Reporting Ledgers}. [Parallel primary (e.g.\ US-GAAP) and secondary (e.g.\ IFRS) ledgers within a single instance with automated conversion journals.]
    \item ISACA. (2024). \textit{AI Shared Responsibility Model: A Governance Framework}.
    \item EU AI Act. (2025). \textit{Governance of High-Risk AI Systems in Financial Markets}.
    \item Deloitte. (2025). \textit{Benchmarks for Global Financial Operations: The Cost of Inefficiency}.
    \item Gartner. (2024). \textit{Transforming Finance through Autonomous Systems}.
    \item Fintech Global. (2025). \textit{The Inefficiency Crisis: Addressing Reconciliation Error Rates in B2B Payments}.
    \item Davidovic, S., \& Tourpe, H. (2026). \textit{How Agentic AI Will Reshape Payments}. IMF Notes, 2026/004. \url{https://www.imf.org/en/publications/imf-notes/issues/2026/04/22/how-agentic-ai-will-reshape-payments-575560}
    \item Google Cloud / Linux Foundation. (2025). \textit{Agent2Agent (A2A) Protocol}. Donated to Linux Foundation June 2025. \url{https://a2a-protocol.org/latest/}
    \item Google / FIDO Alliance. (2025). \textit{Agent Payments Protocol (AP2)}. Donated to FIDO Alliance May 2026. \url{https://ap2-protocol.org/}
    \item World Economic Forum. (2025). \textit{Agentic Commerce and the Future of M2M Transactions}.
    \item OECD. (2023). \textit{Tax Administration 2.0: Structural Digitalization}.
    \item Luciani, C. (2026). \textit{The Accounting Babel: Cross-Border Semantic Fragility}.
    \item G7 Cyber Expert Group. (2026, January). \textit{Roadmap for Coordinating the Transition to Post-Quantum Cryptography in the Financial Sector}. Co-chaired by U.S. Treasury and Bank of England. \url{https://home.treasury.gov/news/press-releases/sb0355}
    \item National Institute of Standards and Technology. (2024, August). \textit{NIST Releases First 3 Finalized Post-Quantum Encryption Standards} (FIPS 203 ML-KEM, FIPS 204 ML-DSA, FIPS 205 SLH-DSA). \url{https://www.nist.gov/news-events/news/2024/08/nist-releases-first-3-finalized-post-quantum-encryption-standards}
    \item Agand, P. (2026). \textit{Neuro-Symbolic Financial Reasoning via Deterministic Fact Ledgers and Adversarial Low-Latency Hallucination Detector}. arXiv:2603.04663. \url{https://arxiv.org/abs/2603.04663} [Universal Fact Ledger; deterministic grounding compresses financial hallucination to 1.2\%.]
    \item Thanh Luong Tuan, \& Abhijit Sanyal. (2026). Ontology-Constrained Neural Reasoning in Enterprise Agentic Systems: A Neurosymbolic Architecture for Domain-Grounded AI Agents. arXiv:2604.00555v5. \url{https://arxiv.org/abs/2604.00555} [Authors listed in full per arXiv; Vietnamese family/given split not asserted. 1,800-run experiment; Vietnamese domains show $2\times$ ontology lift; L0--L5 coupling maturity model; Foundation AgenticOS platform, 22 industry verticals.]
    \item \textit{PHANTOM: A Benchmark for Hallucination Detection in Financial Long-Context QA} (2025). \textit{NeurIPS 2025 Datasets and Benchmarks Track}. \url{https://openreview.net/forum?id=5YQAo0S3Hm} [Authorship to be confirmed at camera-ready.]
    \item Sharma, K., Kumar, P., \& Li, Y. (2025). OG-RAG: Ontology-Grounded Retrieval-Augmented Generation for Large Language Models. \textit{EMNLP 2025}. [$+55\%$ fact recall, $+40\%$ response correctness via ontology-anchored hypergraph retrieval.]
    \item Willard, B. T., \& Louf, R. (2023). Efficient Guided Generation for Large Language Models. arXiv:2307.09702. [Grammar-constrained decoding; syntactic output bounds without fluency loss.]
    \item Luo, L., et al. (2025). Graph-Constrained Reasoning: Faithful Reasoning on Knowledge Graphs with Large Language Models. \textit{ICML 2025}. arXiv:2410.13080. [KG-Trie constrains LLM decoding to valid KG paths at the token level.]
    \item Garcez, A. d., \& Lamb, L. C. (2023). Neurosymbolic AI: The 3rd Wave. \textit{Artificial Intelligence Review} 56: 12387--12406. DOI:10.1007/s10462-023-10448-w.
    \item PricewaterhouseCoopers. (2026, June). \textit{Hyperinflationary Economies as at June 2026}. \url{https://viewpoint.pwc.com/gx/en/pwc/in-briefs/2026/ib_int202610.html}
    \item Ernst \& Young. (2025, November). \textit{IFRS Developments Issue 242: Hyperinflationary Economies}. \url{https://www.ey.com/content/dam/ey-unified-site/ey-com/en-gl/technical/ifrs-technical-resources/documents/ey-gl-developments-242-hyperinflationary-economies-v2-11-2025.pdf}
    \item Spivak, D. I. (2012). Functorial Data Migration. \textit{Information and Computation}. arXiv:1009.1166.
    \item Spivak, D. I., \& Kent, R. E. (2012). Ologs: A Categorical Framework for Knowledge Representation. \textit{PLoS ONE} 7(1): e24274.
    \item Ellerman, D. (2014). On Double-Entry Bookkeeping: The Mathematical Treatment. \textit{Accounting Education} 23(5): 483--501. arXiv:1407.1898.
    \item Robinson, M. (2017). Sheaves are the Canonical Data Structure for Sensor Integration. \textit{Information Fusion} 36: 208--224.
    \item Hansen, J., \& Ghrist, R. (2019). Toward a Spectral Theory of Cellular Sheaves. \textit{Journal of Applied and Computational Topology} 3. arXiv:1808.01513.
    \item Curry, J. (2014). \textit{Sheaves, Cosheaves and Applications}. PhD thesis. arXiv:1303.3255.
    \item Thiéblin, E., Haemmerlé, O., Hernandez, N., \& Trojahn, C. (2020). Survey on Complex Ontology Matching. \textit{Semantic Web} 11(4). DOI:10.3233/SW-190366.
    \item Vauquois, B. (1968). A Survey of Formal Grammars and Algorithms for Recognition and Transformation in Machine Translation. \textit{IFIP Congress}.
    \item Hasić, D., \& Tannier, E. (2019). Gene Tree Species Tree Reconciliation. arXiv:1703.08950.
    \item Mémoli, F. (2011). Gromov--Wasserstein Distances and the Metric Approach to Object Matching. \textit{Foundations of Computational Mathematics} 11(4): 417--487.
    \item Xu, H., Luo, D., Zha, H., \& Carin, L. (2019). Gromov--Wasserstein Learning for Graph Matching and Node Embedding. \textit{ICML}. arXiv:1901.06003.
    \item Cover, T. M. (1965). Geometrical and Statistical Properties of Systems of Linear Inequalities with Applications in Pattern Recognition. \textit{IEEE Trans. Electron. Comput.} EC-14(3): 326--334.
    % --- Labor market section bibliography ---
    \item Bureau of Labor Statistics. (2025a). \textit{Occupational Outlook Handbook: Accountants and Auditors}. U.S. Department of Labor. \url{https://www.bls.gov/ooh/business-and-financial/accountants-and-auditors.htm} [Projected employment change 2024--2034: $+5\%$; 124,200 annual openings; 2.0\% unemployment rate.]
    \item Bureau of Labor Statistics. (2025b). \textit{Occupational Outlook Handbook: Bookkeeping, Accounting, and Auditing Clerks}. U.S. Department of Labor. \url{https://www.bls.gov/ooh/office-and-administrative-support/bookkeeping-accounting-and-auditing-clerks.htm} [Projected employment change 2024--2034: $-6\%$; automation cited as primary cause.]
    \item World Economic Forum. (2025). \textit{Future of Jobs Report 2025}. Geneva: WEF. [Accounting and bookkeeping clerks ranked 7th fastest-declining occupation globally; estimated $-20\%$ by 2030.]
    \item Institute of Chartered Accountants in England and Wales. (2026). \textit{Evolution of Mid-Tier Accountancy Firms}. ICAEW. [Survey of leaders from 35 mid-tier UK firms, February--March 2026: 68\% expect some early-career role reduction; 83\% expect no net role reduction; 71\% report teams moving toward advisory work.]
    \item Mäkelä, M., \& Stephany, F. (2024). AI Complementarity and Labour Market Recomposition: Evidence from 12 Million Job Vacancies. arXiv:2412.19754. Oxford Internet Institute. [AI-exposed roles $2\times$ more likely to require complementary cognitive skills; spillover effect across firm-level job postings.]
    \item PricewaterhouseCoopers. (2025). \textit{Global AI Jobs Barometer}. PwC. [Analysis of close to a billion job ads across six continents: 56\% wage premium for roles requiring AI skills; 7.5\% year-on-year growth in AI-skill postings across sectors.]
    \item Robert Half. (2026). \textit{2026 Accounting and Finance Hiring Guide}. Robert Half International. [CPA-level roles average 73 days to fill, 41\% longer than prior survey period; market described as ``acutely constrained''.]
    \item American Institute of CPAs. (2025). \textit{Trends in the Supply of Accounting Graduates and the Demand for Public Accounting Recruits}. AICPA. [Strong hiring intent across firm sizes; pipeline challenge attributed to reduced CPA-exam sitting rates, not AI displacement.]
    \item Association of Chartered Certified Accountants. (2024). \textit{Strategic Professional Syllabus Update: AI Ethics and Sustainability as Core Competencies, effective 2024}. ACCA. \url{https://www.accaglobal.com/}
    % --- Non-delegability / accountability section bibliography ---
    \item Tucker, J., \& Kapoor, A. (2025, December). \textit{The Accountability Crisis of AI in Assurance: Rethinking Professional Codes of Ethics}. University of Waterloo Centre for Information System Assurance. \url{https://uwaterloo.ca/uwaterloo-centre-for-information-system-assurance/accountability-crisis-ai-assurance-rethinking-professional} [``You cannot fine an algorithm, imprison a neural network, or shame a dataset''; human-in-the-loop as the anchor of professional responsibility.]
    \item Meditari Accountancy Research. (2024). Artificial Intelligence Legal Personality and Accountability: Auditors' Accounts of Capabilities and Challenges for Instrument Boundary. \textit{Meditari Accountancy Research} 32(7): 120--148. Emerald Publishing. [Auditor judgment as a non-delegable core function; against premature AI legal personhood that would dilute human accountability.]
    \item London School of Economics Business Review. (2026, February). \textit{AI Governance Must Move from ``Point-in-Time'' Audits to ``Living'' Compliance}. LSE. \url{https://blogs.lse.ac.uk/businessreview/2026/02/19/ai-governance-must-move-from-point-in-time-audits-to-living-compliance/} [Context integrity as a continuous, not snapshot, governance responsibility.]
    \item Harvard Journal of Law \& Technology. (2025). \textit{Redefining the Standard of Human Oversight for AI Negligence}. \url{https://jolt.law.harvard.edu/digest/redefining-the-standard-of-human-oversight-for-ai-negligence}
    \item Gruber, T. R. (1993). A Translation Approach to Portable Ontology Specifications. \textit{Knowledge Acquisition} 5(2): 199--220. [Canonical definition of ``ontology'' in the information-science sense: a formal, explicit specification of a shared conceptualization.]
    % --- Market-scale / cost-of-fragmentation citations ---
    \item Business Research Industry. (2026, March). \textit{Global Financial Close Software Market Size/Share Worth USD 18.5 Billion by 2034 at an 8.1\% CAGR}. [Financial close software market valued at USD~8.8~billion in 2025.] \url{https://www.globenewswire.com/news-release/2026/03/18/3258374/0/en/}
    \item Market Research (financial consolidation segment). (2025). \textit{Financial Consolidation Software Market Size and Forecast}. [Segment valued at approximately USD~3.2~billion in 2025, CAGR $\approx$9\%.] Figures vary by analyst house; reported here as an order-of-magnitude proxy, not a precise total.
    \item Gartner. (2021). \textit{How to Improve Your Data Quality} / \textit{Data Quality Market Survey}. [Poor data quality costs the average organization an estimated USD~12.9~million per year.] \url{https://www.gartner.com/en/data-analytics/topics/data-quality}
    \item Reta Vortaro (ReVo). \textit{Esperanto dictionary: konto, kontado, kalkuli}. \url{https://vortaro.net} [Source for the Esperanto etymology of ``Kontablo''.]
    \item Genewein, T., Franklin, M., Lerchner, A., Orseau, L., Albanie, S., Bales, A., Wyeth, C., Chan, S., Gabriel, I., Leibo, J.~Z., Dafoe, A., Hutter, M., Graepel, T., \& Legg, S. (2026). \textit{From AGI to ASI}. Google DeepMind. arXiv:2606.12683. [Source of the explicit dual-readership (human/machine) reading-path convention adopted in this paper; cited for document structure, not content.]
\end{itemize}

\end{document}
