\begin{abstract}
\textbf{Kontablo} is an open, graph-based universal accounting ontology anchored to International Financial Reporting Standards (IFRS) and mapped across all 195 sovereign jurisdictions through a universal IFRS-anchored layer, with statutory chart-of-accounts overlays for the 60 jurisdictions that mandate a national chart, providing a deterministic, jurisdiction-agnostic subledger for both multinational ERP consolidation and the Machine-to-Machine (M2M) Agentic Economy.

The global accounting ecosystem remains fragmented across incompatible national standards --- an ``Accounting Babel'' that consumes an estimated 30--40\% of corporate finance teams' time in reconciliation and sustains a multi-billion-dollar financial-close software market. The bottleneck is not the absence of automation but its fragmentation: bespoke point-to-point integrations rather than one shared semantic standard. As AI agents begin to execute financial operations via Agent Payments (AP2) and Agent2Agent (A2A) protocols, they need a semantic execution layer that reporting standards (XBRL) and payment messaging (ISO 20022) do not provide.

Kontablo addresses this gap through three core contributions: (1)~a Level~3 minimum-core taxonomy of 30 universal accounts covering an estimated 94\% of routine transaction volume by posting count ($\sim$99\% with a 34-account extended core), reproducible from a committed benchmark rather than asserted; (2)~a \textbf{Three-Tier Resolution Strategy} combining deterministic exact-code lookups and keyword rules with a confidence-scored semantic AI fallback, bounded by a Deterministic Boundary Library that eliminates the classical accounting hallucination class at the harness level; and (3)~a \textbf{Co-responsibility Architecture (CRA)} pairing every AI mapping proposal with a mandatory human review pathway and an append-only audit trail, keeping legal accountability with the human operator. On a synthetic validation matrix built from the ontology's own committed code sets --- 75 entities, 68 jurisdictions, fourteen languages across five writing systems, four IAS~29 dual-rate hyperinflation stress tests --- the deterministic tiers resolve 97\% of entries, out-of-coverage instruments escalate to human review rather than mismapping, and the CRA quarantines a mechanically wired injected-error catalog. Tier-1 coverage is exercised against primary-source-cited statutory charts in 56 of the 60 statutory-chart jurisdictions (enumerated in the Evaluation section); the run also surfaced four latent code collisions in the standard, since fixed against primary sources. Kontablo is published under the Business Source License~1.1, converting to Apache~2.0 after four years.
\end{abstract}

\vspace{0.8em}
{\small\noindent \textbf{Keywords:} Accounting Standards $\cdot$ Knowledge Graphs $\cdot$ IFRS $\cdot$ Agentic Economy $\cdot$ M2M Transactions $\cdot$ ERP Interoperability $\cdot$ Co-responsibility Architecture $\cdot$ AI Governance $\cdot$ Human-in-the-Loop $\cdot$ Algorithmic Accountability $\cdot$ Sovereign AI $\cdot$ Digital Sovereignty\par}

% --- Front-matter metadata (relocated off page 1 so the abstract stands alone) ---
\vspace{1em}
\noindent\textcolor{clappsAccent}{\rule{\linewidth}{0.4pt}}
\begingroup\scriptsize\raggedright\setlength{\parskip}{0.4em}\par

\noindent\textbf{Author.} Christian Luciani --- Chief Architect, Kontablo Project. Independent Researcher, Cuenca, Ecuador. ORCID:~\href{https://orcid.org/0000-0002-6955-5384}{0000-0002-6955-5384}. Contact:~\href{mailto:cluciani@gmail.com}{cluciani@gmail.com}.

\noindent\textbf{Etymology of ``Kontablo''.} \emph{Kontablo} is a neologism coined for this project. Its roots are Esperanto: \emph{konto} (a financial account) and the cognate \emph{kontado} (accountancy), both tied to \emph{kalkuli} (to calculate, to reckon) (Reta Vortaro, \url{https://vortaro.net}). The form deliberately echoes the Romance \emph{contable}/\emph{contabile} (``accountant'', ``of accounts''), making the term immediately recognizable to Spanish-, Portuguese-, Italian-, and French-speaking readers while remaining neutral in spirit. We read \emph{Kontablo} as ``a system of accounts'': a sustained, calculable accounting discipline rather than a single ledger --- fitting for a universal standard built for a hybrid economy of human and AI agents. (Morphology note: regular Esperanto accountancy derivations use \emph{-ado}/\emph{-eco}; \emph{Kontablo} is a deliberate coinage, not a standard inflection.)

\noindent\textbf{Cite as.} Luciani, C. (2026). \textit{Kontablo: A Graph-Based Universal Accounting Ontology for the M2M Agentic Economy}. Preprint v1.9.2, Zenodo. DOI:~\href{https://doi.org/10.5281/zenodo.20738796}{10.5281/zenodo.20738796}.

\noindent BibTeX: \texttt{@misc\{luciani2026kontablo, author=\{Luciani, Christian\},}\\
\texttt{title=\{Kontablo: A Graph-Based Universal Accounting Ontology for the \{M2M\} Agentic Economy\},}\\
\texttt{year=\{2026\}, howpublished=\{Preprint v1.9.2\}, doi=\{10.5281/zenodo.20738796\}\}}
\endgroup
