\section*{Acknowledgements}

The author thanks Jean González and Luiz González for first making him aware of
the problem Kontablo sets out to solve, and for many valuable discussions;
Gustavo Entrala, David Shapiro, Wes Roth, and Matthew Berman for keeping him abreast of the
accelerating development of artificial intelligence, its risks, and its potential
consequences; and Álvaro Moya for teaching him the rigor and structure of
AI-assisted software engineering. Google's Gemini assisted in the earliest stages
of development, and Anthropic's Claude served as the principal assistant,
development companion, and pair-collaborator throughout the deeper work on the
project. Responsibility for all content, claims, and errors remains the author's
alone (see the AI-use disclosure below).

The author thanks the open-source communities behind ERPNext/Frappe, Odoo, and the IFRS Foundation's public taxonomy releases, whose freely accessible standards made the jurisdiction-mapping analysis tractable. The jurisdictional coverage of Vietnamese (VAS) and Saudi Arabian (SOCPA) standards benefited from AI-assisted bilingual semantic mapping; all resulting mappings were reviewed by the author.

This work was conducted independently. No external funding was received. The Kontablo Project is an initiative of Praxia, Cuenca, Ecuador, currently in incorporation.

% TODO before final submission: add any individual reviewers or domain experts
% who provided feedback on draft versions.

\paragraph{AI use disclosure.} This manuscript was written by the author. Large
language models were used as drafting and editing aids --- restructuring prose,
checking internal consistency, generating critical simulated reviews, and
assisting with the bilingual semantic mappings noted above (all of which were
human-reviewed). All claims, the architecture, the evaluation design, and every
quantitative result are the author's own and are regenerable from committed
commands in the public repository. The dual-readership convention introduced
in \S\ref{sec:summary-instructions} follows Genewein et al.\ (2026); see that
section for attribution.
