\section{Deterministic Logic Rules vs. AI Probability}
\label{sec:deterministic}

\subsection{The Failure of Pure Semantic AI in Accounting}
While modern Large Language Models (LLMs) excel at semantic matching, using them as primary systems of record in financial systems is dangerous. LLM hallucinations in bookkeeping can lead to tax fraud, regulatory penalties, and audit failures. A mapping accuracy of 95\% is acceptable for search queries, but unacceptable for general ledgers where accuracy must be 100\% (or at least all discrepancies must be flagged). An AI proposing that an electricity bill is a ``Prepaid Expense'' instead of an ``Administrative Expense'' introduces systemic errors that compromise financial integrity.

\subsection{The Three-Tier Resolution Strategy}
To achieve high automation without risking hallucinated mappings, Kontablo implements a \textbf{Three-Tier Resolution Strategy} that combines deterministic rules with probabilistic AI fallbacks. When an unmapped localized ERP account is processed, the system routes it through the following pipeline:

\begin{enumerate}
    \item \textbf{Tier 1: Exact Lookup (Direct Mapping):} The engine queries the central Level 3 YAML schema. If the local account code matches a pre-mapped standard code for that jurisdiction (e.g., Mexican SAT code \texttt{101} matches \texttt{asset.current.cash} directly), a link is created. This tier produces an exact match with a confidence score of $1.0$ and requires zero inference latency.
    \item {\sloppy\textbf{Tier 2: Disambiguation Rules (Rule-Based Regex):} If Tier 1 fails, the engine applies keyword and regular expression rules to the account type and name. For example, a Zoho Books account of type \texttt{other\_current\_liability} with a name containing ``VAT'' or ``IVA'' resolves to \texttt{liability.current.vat\_output} (confidence $0.80$--$0.90$).\par}
    \item \textbf{Tier 3: Semantic AI Fallback (Probabilistic Matching):} If both Tier 1 and Tier 2 fail, the engine activates the Semantic Matcher AI. The engine generates a vector embedding of the account's name, type, and description, comparing it against the embeddings of the 30 Level 3 universal accounts. The model outputs a mapping proposal along with a confidence score ($0.50$ to $0.85$). 
\end{enumerate}

\input{figures/fig_three_tier_pipeline}

Figure~\ref{fig:tier_worked_example} traces the same logic on three concrete
inputs --- one resolved at each tier --- to make the distinction between the
deterministic tiers and the single stochastic one concrete.

\input{figures/fig_tier_worked_example}

\subsection{The Deterministic Questions Library}
To prevent LLM mapping proposals from silently corrupting the ledger, all outputs from Tier 3 must pass through a strict library of deterministic boundary rules. Rather than letting the LLM decide a mapping in a single step, the system forces the mapping through a sequence of deterministic questions based on the target node's metadata. 

For example, if the AI proposes mapping a local account ``Petty Cash Fund'' to the Level 3 node \texttt{asset.current.cash}, the rule engine verifies:
\begin{enumerate}
    \item \textbf{Nature Check:} Is the debit/credit nature of the local account consistent with the target node? (e.g. cash is debit).
    \item \textbf{Liquidity Check:} Does the local account represent an asset available at par under 90 days?
    \item \textbf{Boundary Check:} Is the local account a cash equivalent, restricted cash, or digital currency?
\end{enumerate}

If any validation returns a mismatch (e.g., if the AI attempts to map a restricted-use cash account to a standard cash node, or maps a liquid account to a non-current asset node like PPE), the engine overrides the confidence score, setting it to a penalized value of $0.3$, and raises an anomaly flag, forcing manual human-in-the-loop review.

\input{figures/fig_deterministic_checklist}

\subsection{Cross-Jurisdiction Semantic Bridge}
Imagine a multinational with two accountants: one in Mexico who names the cash account ``Caja Revolvente’’ and one in Brazil who uses ``Caixa Geral.’’ To a human ERP manager, these are different. In the Kontablo Graph, both accountants’ local entries hit the same node because they both satisfy the deterministic criteria for `asset.current.cash`.

This allows a corporate headquarters to have a \textbf{Unified Single-target Ledger} where the name of the local account is irrelevant to the consolidated analytical truth. The consolidation occurs at the graph layer, not the string/label layer.

\subsection{The Inverse Parametric Knowledge Effect: Why Non-Anglophone
Jurisdictions Benefit Most}
\label{subsec:ipke}

A non-obvious consequence of ontology-as-constraint is what Luong Tuan \&
Sanyal (2026) term the \textbf{Inverse Parametric Knowledge Effect (IPKE)}:
the value of ontological grounding is \emph{inversely proportional} to the
LLM’s pretraining coverage of the target domain. For well-documented
international concepts -- Basel~III ratios, PCI-DSS levels, IFRS core
terminology -- LLMs already encode sufficient parametric knowledge, and
injecting additional ontological context produces marginal gains (or in
some cases modest regressions due to context displacement). For
enterprise-specific, non-English, or thin-corpus regulatory domains,
ontological grounding is not an optimization but an operational requirement.

This effect has direct implications for Kontablo’s jurisdictional coverage
strategy. The jurisdictions that benefit most from the deterministic
constraint layer are precisely those where pretraining data is structutest
sparse: VAS-Vietnam (government-mandated chart in Vietnamese), SOCPA-Saudi
Arabia (Zakat provisions in Arabic), SYSCOHADA-francophone Africa (OHADA
chart in French with local judicial overlays), and IAS~29 hyperinflationary
contexts (Venezuela, Lebanon, Argentina) where the economic logic departs
from standard IFRS in ways not well-represented in English-language
corpora. The controlled experiment of Luong Tuan \& Sanyal (2026) provides
empirical support: Vietnamese-language financial domains showed $2\times$ the
improvement from ontological grounding relative to English-language domains
($\Delta = +.29$ vs. $+.12$ composite score), replicating across Claude
Sonnet~4, Qwen~2.5~72B, and Gemma~4~26B.

\subsection{Ontology vs. Retrieval-Augmented Generation}
\label{subsec:ontology_vs_rag}

A natural alternative to a formal ontology is a Retrieval-Augmented
Generation (RAG) pipeline over the same regulatory documents. Why is
Kontablo not implemented as a RAG system over IFRS standards, national tax
codes, and SYSCOHADA blueprints?

RAG retrieves; it does not constrain. A RAG system can surface the relevant
section of the Vietnamese VAS standard for a given transaction, but it
cannot \emph{prevent} the LLM from emitting an account code that does not
exist in the client’s actual chart. The critical difference is the locus of
the guarantee:

\begin{itemize}
  \item \textbf{RAG} improves the probability that the model proposes a
  correct mapping. It does not eliminate invalid mappings -- it makes them
  less likely.
  \item \textbf{Ontology-as-constraint} provides a categorical guarantee:
  any UUID that does not exist in the graph is rejected at the harness level
  before reaching the ledger, regardless of the model’s confidence.
\end{itemize}

Luong Tuan \& Sanyal (2026) quantify the distinction: while RAG and ontology
injection achieve comparable scores on lexical recall metrics, ontologies
provide \emph{categorically different} capabilities that document-chunk
retrieval cannot replicate -- hierarchical constraint enforcement, formal
composability across industries, and machine-verifiable output validation
(their Table~5). For Kontablo’s ledger-grade accuracy requirement -- where a
95\% accuracy rate is not acceptable and every mismatch must be flagged --
the categorical guarantee is the operative property, not the average-case
improvement. A system that \emph{usually} maps correctly but \emph{sometimes}
silently posts to a non-existent account is not fit for financial purpose.
