\section{Validation: Mass Consolidation Simulation}
\label{sec:evaluation}

To test the ontology's robustness against the three crises identified in Section 1.1, we developed a mass-consolidation engine and exercised it on \textbf{synthetic} trial balances --- 10 distinct entities across 9 countries representing diverse regulatory environments. The data is simulated, not drawn from production ledgers; what the simulation validates is the mapping and consolidation \emph{architecture}, not field performance.

\subsection{Consolidation Simulation Engine}
The simulation models a multinational holding company consolidating trial balances from 10 distinct operational entities across 9 jurisdictions: Mexico (MX, two entities), Brazil (BR), France (FR), Panama (PA), Ecuador (EC), Colombia (CO), Vietnam (VN), Nigeria (NG), and Saudi Arabia (SA). Each subsidiary ledger is a synthetic trial balance written in the trial-balance export format of one of four source ERPs (ERPNext, Zoho Books, Odoo, SAP Business One) --- the files mirror each ERP's export layout but were never exported from a live ERP instance --- carrying localized charts of accounts (real statutory codes from SAT, SPED-style, PCG, PUC, VAS, SOCPA numbering), varying currencies, and diverging tax treatment. Every entity's trial balance balances exactly (total debits equal total credits in local currency), and the consolidated ledger satisfies the accounting identity $A = L + E + (R - X)$ with escalated lines carried in an explicit suspense bucket. The full artifact --- fixtures, parsers, resolution pipeline, and results --- is committed and regenerable with a single deterministic command (\texttt{scripts/consolidation\_v1\_initial\_run.py}; artifacts in \texttt{research/experiments/consolidation\_v1/}); no API keys or model calls are required. This initial run was subsequently extended into a larger, fully reproducible validation matrix (75 entities, 68 jurisdictions; Section~\ref{subsec:expanded_validation}).

\subsection{The 10-Point Complexity Scale}
To measure the difficulty of mapping local charts of accounts to the universal Level 3 ontology, we established a \textbf{10-point Complexity Scale}. The scale rates jurisdictions based on four primary vectors: (1) alignment with IFRS, (2) strictness of mandatory national charts of accounts, (3) complexity of electronic invoicing and transactional metadata requirements, and (4) currency volatility/inflation treatment. 

Kontablo's full ontology covers all \textbf{195 UN-recognized sovereign states}; the IFRS Foundation's own jurisdiction survey covers 169 jurisdictions (IFRS Foundation, 2025).
Table~\ref{tab:jurisdictions} presents a \textbf{stratified representative sample of 21 jurisdictions} drawn from that full mapping, selected to illustrate the range of complexity tiers from IFRS-pure low-friction environments to IAS~29 hyperinflationary stress cases.
Within this sample, the average complexity score is \textbf{5.6 / 10} (arithmetic mean over the 21 rows; see table), reflecting the deliberate stratification across all difficulty tiers rather than a random draw.

\begin{table}[htbp]
\centering
\small
\resizebox{\textwidth}{!}{%
\begin{tabular}{l|c|l|l}
\hline
\textbf{Jurisdiction} & \textbf{Complexity (1--10)} & \textbf{Primary Complexity Driver} & \textbf{Validation Status} \\ \hline
\multicolumn{4}{l}{\textit{Tier 1--2: IFRS-pure / minimal local divergence}} \\ \hline
United Kingdom (UK) & 2/10 & Full IFRS adoption; common-law flexibility & Fully Mapped \\
Canada (CA)         & 2/10 & IFRS aligned with CPA Canada guidance & Fully Mapped \\
Australia (AU)      & 2/10 & AASB equivalent to IFRS verbatim & Fully Mapped \\
\hline
\multicolumn{4}{l}{\textit{Tier 3--4: Moderate divergence}} \\ \hline
United States (US)  & 3/10 & US-GAAP (FASB ASC); convergence ongoing & Fully Mapped \\
India (IN)          & 4/10 & Ind-AS (IFRS converged, carve-outs apply) & Fully Mapped \\
Germany (DE)        & 4/10 & HGB commercial code statutory divergence & Mapped (IKR Overlay) \\
UAE (AE)            & 4/10 & Federal corporate tax introduced 2023 (MoF) & Fully Mapped \\
\hline
\multicolumn{4}{l}{\textit{Tier 5: Mandatory national chart + digital-tax metadata}} \\ \hline
Japan (JP)          & 5/10 & J-GAAP (ASBJ); mandatory corporate-tax schedules & Fully Mapped \\
Mexico (MX)         & 5/10 & Strict SAT 5-digit catalog tagging & Fully Mapped \\
Colombia (CO)       & 5/10 & Rigid PUC numeric hierarchy & Fully Mapped \\
Panama (PA)         & 5/10 & DGI/SMV dual-standard reporting & Fully Mapped \\
South Africa (ZA)   & 5/10 & SARS guidelines; IFRS adopted via SAICA & Fully Mapped \\
\hline
\multicolumn{4}{l}{\textit{Tier 6: Codified national plans + inflation}} \\ \hline
France (FR)         & 6/10 & Mandatory Plan Comptable G\'{e}n\'{e}ral (PCG) & Mapped (PCG Overlay) \\
Spain (ES)          & 6/10 & PGC 8-group structure; strict numbering & Fully Mapped \\
China (CN)          & 6/10 & CAS standard divergence; VAT reform & Fully Mapped \\
Russia (RU)         & 6/10 & RAS standard; transitioning to IFRS & Mapped (RAS Overlay) \\
Turkey (TR)         & 6/10 & TAS inflation accounting rules (IAS~29 border) & Fully Mapped \\
\hline
\multicolumn{4}{l}{\textit{Tier 7: Complex tax regimes / strict government charts}} \\ \hline
Brazil (BR)         & 7/10 & SPED e-reporting; ICMS/PIS/COFINS tax credits & Fully Mapped \\
Saudi Arabia (SA)   & 7/10 & Zakat religious tax base (SOCPA) & Mapped (Zakat Overlay) \\
Vietnam (VN)        & 7/10 & Strict VAS 2014 government chart mandates & Mapped (VAS Overlay) \\
\hline
\multicolumn{4}{l}{\textit{Tier 10: IAS 29 hyperinflation stress case}} \\ \hline
Venezuela (VE)      & 10/10 & IAS~29 hyperinflation; dual-currency parallel rates & Stress Tested \\
\hline
\end{tabular}%
}
\caption{Stratified representative sample (21 of 195 mapped jurisdictions) illustrating the full complexity-tier range of the Kontablo ontology. Complexity scores reflect: IFRS alignment, mandatory national chart strictness, digital-tax metadata requirements, and currency/inflation treatment. Mean complexity for this sample: 5.6/10. US-GAAP source: FASB ASC (fasb.org); Ind-AS source: MCA notification 2015 (mca.gov.in); J-GAAP source: ASBJ (asb.or.jp). Full 195-jurisdiction dataset available in the project repository.}
\label{tab:jurisdictions}
\end{table}

\subsection{Coverage and Data Quality}
The AI-augmented graph successfully mapped 100\% of the sample transaction sets to Kontablo's core `asset.current.cash` and `asset.current.receivables` universal IDs. By dynamically applying local-to-target foreign exchange conversions and semantic rate overrides, the engine demonstrated that Kontablo can seamlessly aggregate localized financial ledgers into a unified global balance sheet.
A separate, reproducible transaction-frequency benchmark
(\texttt{scripts/coverage\_benchmark.py}, dataset and provenance in
\texttt{research/coverage\_benchmark/}) estimates that the 30 Level~3 accounts
cover \textbf{$\sim$94\% of routine transaction volume by posting-line count}
(and $\sim$87\% counted by whole transaction). We are explicit about the
evidentiary status of this figure: it is a \emph{model-based estimate} computed
over a transparently-parameterized synthetic frequency distribution, not a
measurement of a proprietary ledger corpus. The distribution's structure is
grounded in the standard double-entry ``special journals'' taxonomy (Weygandt,
Kieso \& Kimmel; Horngren et al.), under which the bulk of postings flow through
four high-volume journals (Sales, Cash Receipts, Purchases, Cash Payments) whose
transaction types map entirely onto core nodes, while the General Journal carries
only infrequent adjusting entries. The committed artifact replaces an earlier,
unbacked appeal to a large but uncommitted set of ledger exports; substituting a
licensed real-ledger corpus is a documented next step that leaves the script
unchanged.
The substantive finding is robust to the exact weights: extreme ledger
granularity in national standards (often hundreds of accounts) is mostly reserved
for rare statistical or edge-case events.

\subsection{Coverage Expansion and the Principled Residual}
\label{subsec:coverage_expansion}

The $\sim$94\% figure is a coverage-by-volume measure of a \emph{minimum core}
of 30 nodes, not a ceiling. Because transaction frequency follows a long-tail
(Pareto-type) distribution --- a small number of account types carry the bulk
of posting volume, while a long tail of specialized types occurs rarely ---
coverage is raised not by redesigning the architecture but by \textbf{extending
the ontology with additional Level~3 nodes for the next tier of transaction
types}. The same benchmark identifies these next-tier types directly from the
residual: in descending volume order they are payroll net-pay and
statutory-withholding liabilities, contract/deferred revenue (IFRS~15), tax
withheld at source on income, and non-trade receivables. Adding exactly these
\textbf{four} nodes --- a 34-account \emph{extended core}, kept separate from the
frozen 30-account minimum core --- lifts modeled volume coverage from
$\sim$94\% to \textbf{$\sim$99\%} ($\sim$97\% by whole transaction). This both
confirms and sharpens the earlier rough estimate (which guessed ``a few dozen''
nodes): in this profile the bulk of the residual collapses onto a handful of
high-frequency payroll- and tax-related liabilities. The final percentage points
correspond to genuinely idiosyncratic, jurisdiction-specific, or novel
instruments, which we deliberately leave to escalation rather than core growth.

We deliberately do \textbf{not} claim, or target, 100\% coverage. Two reasons,
one principled and one architectural. \emph{Principled:} a universal standard
that asserted total coverage would be epistemically overconfident --- new
instruments (tokenised assets, CBDC settlement, novel Islamic-finance
structures) and new jurisdictional rules continuously regenerate the tail, so a
closed claim of completeness would be false the moment it was made. A residual
uncertainty band is the honest representation of an open, evolving domain.
\emph{Architectural:} Kontablo already has a designated mechanism for the
residual --- the semantic coverage boundary (Section~\ref{subsec:residual_error})
and the Co-responsibility Architecture (Section~\ref{sec:harness}), which route
out-of-coverage transactions to human review rather than forcing a
low-confidence posting. Coverage expansion shrinks the residual; the CRA
guarantees that whatever residual remains is escalated and made observable, not
silently mismapped. The objective is therefore not coverage~$=100\%$ but
coverage~$\to$~high with a \emph{bounded, instrumented, human-resolved}
remainder.

This raises a governance question the benchmark answers directly: \emph{under
what criteria may the core grow, and when must it not?} We codify the answer as
an explicit admission policy (ADR-015). The ontology is structured in three
layers --- a \textbf{frozen} 30-node minimum core (the citable Pareto headline),
a \textbf{bounded} extended core that grows only under the policy, and an
open-ended residual handled by overlays and CRA escalation that is \emph{never}
forced into the core. A transaction type earns a core node only if it
simultaneously (i)~captures material, \emph{measured} posting volume (the
benchmark is the gate, with a working threshold near $0.5\%$ of total volume),
(ii)~recurs across many jurisdictions under a clean IFRS anchor (otherwise it is
a local overlay, not a universal node), (iii)~admits an unambiguous deterministic
boundary --- a fixed nature, statement class and IFRS tag with no collision ---
(iv)~is not already decomposable into existing nodes, and (v)~is a mature, stable
instrument. Because posting volume is Pareto-distributed, this gate is
\textbf{self-limiting}: after the extended core, almost every further candidate
falls below the volume threshold, so the core converges on a small number of
dozens of nodes rather than the hundreds found in national charts. Growth by
intuition or by request is thereby replaced with growth by measured evidence,
and the standing tension between a small, elegant core and broad coverage is
resolved without compromising either.

\subsection{Expanded Multi-Jurisdiction Validation}
\label{subsec:expanded_validation}

\paragraph{What ``195 jurisdictions'' means.} Coverage is recorded in an
explicit manifest of all 195 sovereign states --- the 193 UN member states plus
the two UN observer states (Holy See, State of Palestine) --- in
\texttt{core/schemas/jurisdiction\_coverage.yaml}. The claim is precise and
two-layered. \emph{Every} one of the 195 is covered by Kontablo's universal
IFRS-anchored layer, because each Level-3 node carries an \texttt{ifrs\_tag};
this is the mapping basis for the \textbf{135} jurisdictions that do not mandate
a national numeric chart of accounts (typically common-law / IFRS-direct
regimes --- including IFRS-converged systems such as Japan, South Korea,
Indonesia, Thailand and Switzerland, and IFRS-adopting CIS states such as
Armenia, Georgia, Azerbaijan, Kyrgyzstan and Tajikistan, that prescribe
reporting rules but no legally mandated numeric chart --- where an entity
designs its own chart and there is no statutory code to map). The remaining
\textbf{60} jurisdictions impose a statutory chart and additionally receive a
local-code overlay; for \textbf{56} of these we populate
and exercise primary-source-cited Tier-1 code sets, including the SYSCOHADA
chart shared identically by all 17 OHADA member states (one verified chart, 17
jurisdictions); the Spanish PGC (RD~1514/2007), the Portuguese SNC (Portaria
218/2015), the Peruvian PCGE, the Chinese CAS (MOF) and the Belgian PCMN
(CNC/CBN); the Moroccan CGNC, the Algerian SCF and the Romanian chart
(OMFP~1802/2014); the Czech (Vyhláška 500/2002), Slovak, Hungarian (Act~C/2000)
and Bulgarian national charts; the Ukrainian (Order~291), Kazakh, Tunisian
(PCT~1997) and Belarusian charts; the Serbian, Croatian (RRiF), Slovenian and
Austrian (EKR) charts; the Luxembourg PCN and the French PCG (also applied in
Monaco); the Moldovan and Greek (EGLS/ELP) charts; and single-country charts
such as Panama's (11 cited codes) and Mexico's SAT. The four statutory-chart
jurisdictions not yet carrying Tier-1 codes (Poland, Uzbekistan, Turkmenistan,
Laos) mandate a chart whose account numbers we could not confirm from a primary
source --- Poland's are not legally fixed --- and are left Tier-1-pending rather
than transcribed speculatively (boundary B1).

\paragraph{Boundary conditions of the coverage claim.} To keep the 195 figure
honest rather than rhetorical, the manifest is governed by an explicit
construction protocol, recorded in its metadata: \emph{(B1)} no fabricated
codes --- a jurisdiction is marked Tier-1 only when a primary-source-cited code
set exists, and a deterministic filter (a real statutory code must contain at
least one digit) excludes descriptive text placeholders such as ``Cash'' or
``Vorsteuer'' from the Tier-1 index (14 such placeholders were demoted, and the
affected common-law jurisdictions fall back to the IFRS-tag basis); \emph{(B2)} two distinct claims are reported separately ---
\texttt{mapping\_mode=statutory\_chart} (a mandated chart \emph{exists}) versus
\texttt{tier1\_codes\_available} (we hold \emph{verified} codes), so ``statutory
without codes'' never masquerades as Tier-1 coverage; \emph{(B3)} IFRS-direct is
a genuine coverage mode, not a gap, since the IFRS tag is the complete mapping
basis where no local code exists; \emph{(B4)} multi-state multipliers are used
only where a chart is genuinely shared (SYSCOHADA across 17 states; the EU,
having no shared chart, is covered per country); \emph{(B5)} jurisdictions that
apply IFRS without a mandated chart (e.g.\ Chile, Uruguay, Costa Rica) are left
IFRS-direct rather than asserting a chart that does not exist; \emph{(B6)} IFRS
adoption status derives from the IFRS Foundation profiles, with unprofiled
states marked as such. The coverage claim is thus the conjunction of a
universal IFRS-anchored layer and a verifiable, conservatively-bounded statutory
overlay --- not an assertion of 195 transcribed national charts.

To test the pipeline beyond the initial proof, we ran a broader, fully
reproducible consolidation (script \texttt{scripts/mass\_consolidation\_v2.py};
artifacts in \texttt{research/experiments/consolidation\_v2/}). The trial
balances are \emph{generated directly from the ontology and chart families
themselves}: for every jurisdiction with real \texttt{local\_codes} (inline or
via a chart family), an entity is constructed from those exact codes, so Tier-1
exact lookup is exercised against genuine mapping data rather than hand-authored
examples. The run
processes \textbf{366 local account entries from 75 entities across 68
jurisdictions}, spanning all five statement classes (assets, liabilities,
equity, revenue, expense) and fourteen languages across five writing systems
(Latin: Spanish, Portuguese, French, Italian, Polish, Indonesian, Vietnamese,
German, Turkish, English; Greek; Cyrillic: Russian; Arabic; Hangul: Korean),
and populates \textbf{25 of the 30 universal Level-3 nodes}. To keep
the run deterministic and reproducible, the semantic AI fallback (Tier~3) is
\emph{not} exercised; instead, any entry the deterministic tiers cannot resolve
is counted as \textbf{escalated} (routed to human review by the CRA). The run is
therefore a test of pipeline correctness and boundary behaviour on a controlled
dataset, not an out-of-sample accuracy benchmark.

\begin{table}[htbp]
\centering
\small
\begin{tabular}{l r}
\hline
\textbf{Metric} & \textbf{Result} \\ \hline
Sovereign states in coverage manifest & 195 \\
\quad universal IFRS-anchor layer & 195 \\
\quad statutory-chart overlay & 60 \\
\quad IFRS-direct (mapping via IFRS tag) & 135 \\
\quad Tier-1 code sets populated (cited) & 56 / 60 statutory \\
\hline
Entities consolidated & 75 \\
Jurisdictions exercised in run & 68 \\
IAS~29 hyperinflation dual-rate cases & 4 (VE, AR, LB, TR) \\
Local account entries processed & 366 \\
\hline
Tier~1 (exact local-code lookup) & 310 \quad(84.7\%) \\
Tier~2 (deterministic keyword rules) & 46 \quad(12.6\%) \\
\textbf{Deterministic coverage (Tier~1+2)} & \textbf{97.3\%} \\
Escalated to human via CRA & 4 \quad(1.1\%) \\
\hline
Injected-error catalog detected \& quarantined by CRA & 8 / 8 \\
\quad distinct invariant classes exercised & 5 \\
\quad (nature, liquidity, statement, VAT direction, equity/liability) & 12 flags \\
Descriptive placeholders excluded from Tier-1 (B1) & 14 \\
Ontology code-collisions: detected $\to$ corrected $\to$ remaining & 4 $\to$ 0 \\
\hline
Distinct Kontablo nodes populated & 25 / 30 \\
Consolidated total assets (FX-normalized) & USD 14{,}746{,}038 \\
\hline
\end{tabular}
\caption{195-jurisdiction coverage manifest and the deterministic validation run
(75 entities, 68 jurisdictions), with Tier-1 trial balances generated from the
ontology's own \texttt{local\_codes} and statutory chart families (e.g.,
SYSCOHADA for the 17 OHADA states). Coverage is measured by entry count; FX
normalization to USD uses simulated 2026 rates, with parallel-rate overrides for
hyperinflationary entities. The hyperinflation cases exercise IAS~29's
dual-rate (official vs.\ parallel) translation divergence only; CPI restatement
of non-monetary items, which IAS~29 requires before translation, is not
exercised by this harness.}
\label{tab:expanded_validation}
\end{table}

Three results are worth emphasising. First, the \textbf{four escalated entries are
all genuine coverage-boundary cases}, not failures: three are v0.3-roadmap
instrument types not yet ratified in the operational ontology (cryptocurrency
holdings, carbon credits, and a Zakat provision --- all carried in the YAML
\texttt{pending\_accounts} block without a confirmed \texttt{nature}), and one
is a jurisdiction-specific German tax-driven reserve
(\textit{Sonderposten mit R\"ucklageanteil}) with no clean IFRS counterpart.
The engine \emph{escalated rather than mismapped} each of them --- the
empirical instantiation of the principled residual of
Section~\ref{subsec:coverage_expansion}. Second, the Co-responsibility
Architecture was tested against a deliberately injected error catalog spanning
\textbf{five deterministic accounting invariants} --- debit/credit nature,
the liquidity boundary (liquid asset to non-current), statement class (a P\&L
name forced onto a balance-sheet node and vice versa), VAT direction (input
forced onto the output node and vice versa), and equity/liability
classification. It \textbf{detected and quarantined all eight injected errors}
(twelve flags in total), holding them for human review rather than posting them.
No injected error passed silently, and none reached the consolidated ledger.
The scope of this result must be stated precisely: the nature and liquidity
checks (invariants one and two) run unconditionally on every resolved mapping,
but the statement-class, VAT-direction, and equity/liability checks run only
on entries pre-labeled as upstream proposals (\texttt{forced\_id}), and each
injected name was written to contain vocabulary the corresponding check
matches. The catalog therefore demonstrates that each invariant class is
\emph{mechanically wired end-to-end} --- a flagged entry is provably
quarantined before posting --- not a detection-rate measurement; recall on
plausible-but-wrong proposals whose names evade the keyword vocabulary is
unmeasured and is future work. Third, and most consequential
for the development of the standard, the run \textbf{closed a corrective loop on
the ontology itself}. It surfaced four latent local-code collisions --- a single
jurisdiction code mapped to two different Kontablo nodes (MX \texttt{509}
$\to$ interest/fx-loss; RU \texttt{68.2} $\to$ input/output VAT; VN \texttt{311}
$\to$ short-/long-term debt; VN \texttt{635} $\to$ interest/fx-loss). Each was
then corrected against the primary national standard (SAT \emph{c\'odigo
agrupador} Anexo~24 for MX; the Russian chart Order~94n, which separates input
VAT in account~19 from VAT settlements in 68-02; and Vietnam's Circular
200/2014, which assigns long-term borrowings to account~341). The VN \texttt{635}
case was recognised not as a typo but as a genuine \emph{granularity mismatch} ---
VAS aggregates interest and FX losses under one statutory code --- and resolved
by mapping the code to its dominant leaf and documenting the aggregation rather
than fabricating a spurious distinct code. After correction, re-validation
reported \textbf{zero} collisions. This is a concrete instance of Kontablo's
central thesis operating on Kontablo: a machine-verifiable standard is a
self-checking one. The harness functions not only as a mapping engine but as a
continuous data-quality check on the standard it implements; the full
find$\to$fix$\to$re-validate record is archived with the run artifacts.

\paragraph{Limitation: what the 97.3\% figure does and does not measure.}
The deterministic-coverage figure is an \emph{internal-consistency and
data-quality result}, not a field-accuracy measurement. The Tier-1 entries
are generated by iterating over the ontology's own committed
\texttt{local\_codes} and chart families and are then resolved against an
index built from the same data, so Tier-1 success is guaranteed by
construction except where the data itself is defective --- which is precisely
what makes the run a self-check: the four code collisions it surfaced are
real defects it caught in its own answer key. The Tier-2 entries are
author-written names matched against author-written multilingual rules, and
the four escalations are deliberately out-of-vocabulary frontier instruments.
None of this exercises the noise of real ledger exports --- abbreviations
(``A/R Trade''), department-suffixed codes (``1100-01-DEPT7''), or
misspellings --- and Tier-2 substring matching should be expected to degrade
on such input. A perturbation experiment (typos, prefixed codes,
abbreviations) over the committed fixtures is planned future work; until it
is committed, the honest reading of 97.3\% is ``the pipeline mechanics and
the ontology's code data are mutually consistent,'' not ``97\% of real-world
ledger lines resolve deterministically.''

\subsection{Non-Anglophone Jurisdictions and the Value of Deterministic Grounding}

The initial run exposes a pattern across the non-Anglophone jurisdictions
rated high on the complexity rubric of Table~\ref{tab:jurisdictions}
(Vietnam 7/10, Saudi Arabia 7/10) that aligns with the \textbf{Inverse
Parametric Knowledge Effect (IPKE)} identified in independent empirical work
(Luong Tuan \& Sanyal, 2026; Section~\ref{subsec:ipke}): ontological grounding
adds most value precisely where LLM pretraining coverage of the target domain
is structurally thinnest.

The committed run (\texttt{research/experiments/consolidation\_v1/results.json})
shows the two sides of this effect. For Saudi Arabia (SOCPA with Zakat
overlay), the deterministic tiers resolve only 5 of 12 lines --- one
ontology-carried SOCPA code at Tier~1, four Arabic names matched by the
multilingual Tier~2 rules --- and the remaining 7 lines escalate to human
review, the highest escalation rate in the run. (In these reproducible runs,
lines the deterministic tiers cannot resolve are \emph{escalated} rather than
passed to the Tier~3 semantic fallback, which would require model calls; in a
deployment, these escalations are exactly the lines that Tier~3 would
receive.) For Vietnam, by contrast, the ontology carries the VAS Circular~200
statutory code set, and 12 of 13 lines resolve at Tier~1 --- committed
deterministic grounding substitutes entirely for the parametric knowledge the
model lacks. The contrast is the point: where neither committed codes nor
keyword rules cover a jurisdiction, the residual routes to semantic inference
or to a human, and without the Deterministic Boundary Library's hard rejection
of UUID non-members those would be the lines most likely to produce silent
misclassifications. The deterministic constraint layer is therefore not
merely a quality enhancement; it is an operational requirement for
jurisdictions where the model's parametric knowledge is sparse.

This observation generalises. Of the 195 jurisdictions mapped, those with
non-Latin charts and thin English-language regulatory documentation
(SYSCOHADA francophone Africa, Korean K-GAAP, Ukrainian P(S)BO, Arabic SOCPA,
Vietnamese VAS) are precisely the jurisdictions where Kontablo's constraint
layer provides the highest marginal safety guarantee over an unconstrained RAG
or pure-LLM approach. The universality of coverage, argued structurally in
Appendix~D, is reinforced empirically by the finding that the hardest
jurisdictions to map are also the ones where hard constraints matter most.
