\section{Glossary of coined and key terms}
\label{sec:glossary}

The following terms are defined once here so that they are unambiguous on first
encounter and attributable when reused. Coinages specific to this work are marked
\textsuperscript{\dag}; the remainder are established terms used in a specific
sense by Kontablo.

\begin{description}
  \item[Accounting Babel] The persistent fragmentation of the global accounting
    ecosystem across mutually incompatible national chart-of-accounts standards,
    by analogy to the biblical confusion of tongues. The condition Kontablo is
    designed to bridge.
  \item[Co-responsibility Architecture (CRA)\textsuperscript{\dag}] The governance
    pattern in which every AI mapping proposal is paired with a mandatory human
    review pathway and an append-only audit trail, keeping legal accountability
    with the human operator, who is always the legal principal. High model
    confidence changes the friction of review, never its presence.
  \item[Deterministic Boundary Library\textsuperscript{\dag}] The bounded set of
    deterministic rules and constraints that fences the semantic AI fallback,
    eliminating the classical accounting hallucination class at the harness level
    by making it impossible to propose an account UUID that does not exist in the
    graph.
  \item[Financial Integration Velocity (FIV)\textsuperscript{\dag}] A proposed
    measure of how quickly two previously separate financial systems can be made
    semantically interoperable, used to frame the M\&A and consolidation impact of
    a shared ontology.
  \item[Harness] The deterministic software envelope around the language model:
    the three-tier resolver, the Deterministic Boundary Library, and the ontology
    loaders. The paper's central model--harness distinction holds that the locus
    of residual error relocates from the model to the harness's semantic coverage
    boundary.
  \item[Inverse Parametric Knowledge Effect\textsuperscript{\dag}] The observation
    that deterministic grounding yields the largest benefit precisely where a
    language model's parametric knowledge is weakest --- non-Anglophone and
    lower-resource jurisdictions --- because the ontology supplies what the model
    lacks.
  \item[Level 3 Minimum Core] The taxonomy of 30 universal accounts estimated to
    cover \textasciitilde94\% of routine transaction volume by posting count
    (\textasciitilde99\% with a 34-account extended core), reproducible from a
    committed benchmark.
  \item[Pacioli Group] The accounting-native algebraic anchor for the
    double-entry invariant (debits equal credits), named for Luca Pacioli; used in
    the mathematical-foundations treatment of the translation mechanism.
  \item[Three-Tier Resolution Strategy\textsuperscript{\dag}] The mapping
    pipeline combining (1)~deterministic exact-code lookup, (2)~keyword /
    disambiguation rules, and (3)~a confidence-scored semantic AI fallback ---
    in strict determinism-first order, so stochastic inference is reached only
    when deterministic resolution is exhausted.
  \item[Tree-to-Graph Universal Bridge\textsuperscript{\dag}] The core
    architectural move of representing accounts as nodes in a multi-dimensional
    property graph rather than a rigid tree, enabling lossless round-trip
    linearization back to any target tree (e.g.\ an ERP export or an IFRS
    statement).
\end{description}
