% ============================================================
% Section: The Kontablo Agent — Harness Architecture and
%          the Locus of Error
% Placement: after deterministic_logic.tex, before agentic_economy.tex
% Status: new section added Phase 2.5 pre-publication pass
% Epistemic notes:
%   - Claims marked [VERIFICAR] require a literature check before
%     final submission. Do not remove the marker; replace it with
%     a confirmed citation or remove the claim.
%   - Claims marked [PROPUESTA ORIGINAL] are design contributions
%     without prior literature support identified. They must be
%     defended as such in peer review, not presented as established
%     facts.
% ============================================================

\section{The Kontablo Agent: Harness Architecture and the Locus of Error}
\label{sec:harness}

\subsection{The Model--Harness Distinction}
\label{subsec:model_harness}

A common framing of AI reliability in critical systems asks:
\emph{``How do we prevent the LLM from hallucinating?''}
This question, while intuitive, addresses the wrong architectural layer.
It locates the reliability problem inside the model, where the design levers
available to system architects are limited and where the state of the art
offers no near-term guarantees suitable for ledger-grade accuracy.

A more productive framing asks:
\emph{``How do we design the system around the model so that the
decision space the model can influence is constrained to valid outputs,
and anything outside coverage escalates to a human rather than being
confabulated?''} This reframe is the starting point of harness-centric
agent design.

The \textbf{model} is the stochastic component: the Large Language Model that
performs semantic interpretation, natural-language matching, and
open-ended reasoning over unstructured input. The \textbf{harness} is the
deterministic scaffold surrounding it: the tool definitions and schemas
that structure its output, the ontological constraints that bound the valid
output space, the validation layers that reject structurally invalid proposals
before they reach the ledger, and the escalation protocols that route
out-of-coverage cases to human review.

The core practical insight is that the reliability of an agentic system in a
critical domain is determined far more by the harness than by the model's
intrinsic accuracy. This has consequences for where engineering effort should
be concentrated and where reliability guarantees can realistically be made.
Recent work on agent scaffolding and tool-augmented language models
substantiates this view. Willard \& Louf (2023) demonstrate that
grammar-constrained generation -- where a formal grammar bounds the output
token space -- produces structurally valid outputs without sacrificing fluency.
Luong Tuan \& Sanyal (2026) extend this to enterprise multi-industry deployment
across 1,800 controlled runs, showing that ontology-coupled agents significantly
outperform ungrounded baselines on metric accuracy ($p < .001$, Kendall's $W =
.460$) and role consistency ($p < .001$, $W = .614$) across three model
architectures. Critically, their neurosymbolic coupling maturity taxonomy
formalizes the distinction between systems that constrain \emph{inputs} (context
injection, tool discovery, governance thresholds) and systems that validate
\emph{outputs} -- a distinction we return to in Section~\ref{subsec:coupling_maturity}.

\subsection{Ontology-as-Constraint: Eliminating the Classical Accounting
Hallucination Class}
\label{subsec:ontology_constraint}

In the absence of a harness, an LLM deployed for transaction classification
can hallucinate in an accounting-specific way: it may propose an account code
or name that \emph{sounds} plausible but does not correspond to any account in
the client's chart of accounts, or it may classify a transaction to the wrong
node with high expressed confidence and no observable uncertainty signal.

The asymmetry with information retrieval is severe. A 95\% classification
accuracy is acceptable in search; in a general ledger, the 5\% of
misclassified transactions introduces systematic distortions that compound
across reporting periods, survive into audited financial statements, and --
in the presence of the Three-Tier Resolution pipeline -- would do so
\emph{silently}, with no inconsistency flag raised, because an unconstrained
model emitting a plausible-looking but wrong UUID would pass surface validation.

Kontablo addresses this through what we term \textbf{ontology-as-constraint}:
the agent's tool interface (exposed via the MCP server) accepts only Kontablo
Level 3 UUIDs as valid account identifiers. The set of valid UUIDs is finite,
enumerable, and maintained in the ontology graph. The harness rejects any
proposed UUID that does not exist in the graph at validation time, before
the value reaches the ledger write path.

This mechanism is analogous to grammar-constrained or schema-constrained
generation in the structured output literature (Willard \& Louf, 2023; Luo
et al., 2025), where the valid output space is defined by a formal grammar
and outputs violating that grammar are rejected at inference time. In
Kontablo's case the ``grammar'' is the ontology itself: the closed set of
valid account UUIDs, each carrying its nature (debit/credit), IFRS tag, and
deterministic boundary predicates from the Deterministic Boundary Library.

The consequence is that the \emph{classical accounting hallucination class} --
the emission of arbitrary account identifiers that do not exist in the
ontology -- is architecturally eliminated at the harness level.
This is not a probabilistic reduction in hallucination rate; it is a
categorical elimination of a specific error class within the constrained
output space. The model can still be wrong within the set of valid UUIDs
(mapping a transaction to \texttt{expense.admin} when \texttt{expense.cogs}
is correct), but it cannot emit a non-existent account.

\textbf{Implementation dependency:} This guarantee holds only if the harness
enforcement is strict and complete -- no code path allows a UUID to reach the
ledger write step without a graph-membership check. This is a design
requirement, not an automatic consequence of using an LLM with tool
definitions. Any deployment that bypasses the validation layer for performance
or convenience reasons loses this guarantee.

\subsection{The Relocation of Residual Error}
\label{subsec:residual_error}

Eliminating the classical hallucination class does not eliminate error; it
\emph{relocates} it. The residual error population, after ontology-as-constraint
is applied, concentrates at a specific and identifiable location: the
\textbf{boundary of semantic coverage}.

\input{figures/fig_locus_of_error}

Errors at the semantic coverage boundary take three forms:

\begin{enumerate}
    \item \textbf{Unmapped transaction types.} A transaction type not yet
    represented in the ontology has no valid UUID. The Three-Tier Resolution
    Pipeline routes it to Tier 3 (semantic AI fallback); if confidence remains
    below the threshold (Section~6), it is flagged for human review rather than
    forcing a low-confidence posting. The error is not confabulated -- it is
    escalated and made observable.

    \item \textbf{Jurisdictional edge cases.} A jurisdictional rule not yet
    encoded in the Deterministic Boundary Library may allow a semantically
    plausible UUID to pass boundary validation while violating an implicit
    local constraint. These errors are attributable to incomplete rule coverage,
    not to model hallucination. They are also correctable through ordinary
    ontology maintenance as cases are discovered.

    \item \textbf{Semantic ambiguity within coverage.} Some transactions
    satisfy the boundary predicates of more than one Level 3 node. In these
    cases Tier 3 may produce a defensible but suboptimal mapping. The
    Co-responsibility Architecture (Section~10) manages this residual by
    requiring human review for low-confidence Tier 3 outputs.
\end{enumerate}

This relocation is architecturally significant because boundary errors are
\textbf{observable}: they manifest as flagged inconsistencies, low-confidence
outputs, and unmapped-transaction queues -- all instrumentable signals.
Stochastic interior errors from an unconstrained model are, by contrast,
\textbf{silent}: a confident-but-wrong classification produces no observable
artifact beyond the wrong posting itself, invisible to the audit trail unless
a human catches it retrospectively.

This relocation framing is corroborated by independent work.
Luong Tuan \& Sanyal (2026) describe the equivalent phenomenon in their
neurosymbolic enterprise architecture as the \emph{inverse parametric
knowledge effect}: the value of ontological grounding is highest precisely at
the coverage boundary, where LLM parametric knowledge is weakest (Vietnamese
financial regulation: $\Delta = +.29$; English financial domains: $\Delta =
+.12$). A system well-grounded in the core of the ontology reveals its
failure mode at the periphery -- which is exactly the observable error
behavior the CRA is designed to capture, not suppress.
The neuro-symbolic financial reasoning architecture of [arXiv:2603.04663]
(2026) similarly identifies that shifting from probabilistic retrieval to
deterministic grounding compresses financial hallucination to 1.2\%, with
residual error concentrating at domain boundaries -- confirming the relocation
pattern as a general property of constrained-output financial systems, not an
artifact of Kontablo's specific design.

\subsection{The Engineering Reframe: From Intractable to Tractable}
\label{subsec:reframe}

The classical framing -- ``how do we stop the model from hallucinating?'' --
is not only intractable as stated, but directs engineering effort at the wrong
target. Hallucination in the general sense is an active area of model research
with no near-term resolution that guarantees the classification accuracy a
ledger requires across all inputs and distribution shifts.

The productive reframe, which Kontablo's architecture instantiates, is:

\begin{quote}
\emph{How do we design the harness so that (a) the output space the agent
can emit is bounded by the ontology, making confabulation of non-existent
accounts architecturally impossible; (b) decisions within the bounded space
that can be resolved by deterministic logic are resolved without inference;
and (c) cases outside coverage or below confidence threshold are escalated
to a human co-signer rather than silently committed?}
\end{quote}

This is a tractable engineering problem with identifiable components:

\begin{itemize}
    \item Schema-constrained tool interfaces that enforce UUID membership
    \item The Three-Tier Resolution Pipeline (Section~6), which maximizes
    the proportion of decisions resolved deterministically before reaching
    the model
    \item The Deterministic Boundary Library (Section~6.3), which validates
    even Tier 3 proposals against ontological predicates before ledger write
    \item The Co-responsibility Architecture (Section~10), which governs
    the out-of-coverage escalation path
\end{itemize}

Kontablo's architectural principle~5 -- ``determinism over stochasticity
wherever movable'' (ADR-009) -- formalizes this reframe as a standing design
driver: whenever a decision can be migrated from stochastic inference to
deterministic logic, it must be. The harness is the system-level
implementation of this principle. The ontology-as-constraint is its primary
instrument.

\subsection{The Co-responsibility Architecture as Residual Error
Management}
\label{subsec:cra_as_error_management}

The Co-responsibility Architecture (ADR-008, Section~10) serves two functions
(Section~\ref{subsec:non_delegability}): it is the foundational locus of
non-delegable legal accountability for every transaction, and --- the aspect
this subsection isolates --- it is the \textbf{residual error management
layer}. The discussion here concerns only the second, reactive function; it
presupposes, and does not replace, the foundational accountability the human
carries on every record regardless of whether it is flagged.

Where the three-tier pipeline and the Deterministic Boundary Library resolve
decisions within coverage, the CRA governs what happens at and beyond the
coverage boundary. When the harness cannot deterministically resolve a
transaction -- because no Tier 1 or Tier 2 rule matches, and Tier 3 returns
confidence below threshold -- the CRA routes the transaction to human review,
enforces justification logging, and records the outcome in an append-only
audit trail (the specification calls for cryptographic sealing of this
trail; v0.x implements append-only logging).

Within this reactive scope, the human accountant is not a rubber-stamp
approver of AI output. For the cases that reach review, they are the
\textbf{resolver of the residual error population}: cases genuinely beyond
automated coverage that require human judgment. The CRA's inconsistency
flagging mechanism ensures that every such resolution is recorded and
attributable, operationalizing the ISACA shared-responsibility model and the
EU AI Act's requirements for high-risk AI systems in financial contexts
(ISACA, 2024; EU AI Act, 2025). This is precisely the human-in-the-loop
anchoring that recent work on the AI assurance accountability gap identifies
as the mechanism preserving professional responsibility when autonomous
systems cannot bear it (Tucker \& Kapoor, 2025).

This framing implies a property of the system over time that applies
\textbf{only to the reactive function}: as the ontology expands -- more
jurisdictions mapped, more transaction types covered, more boundary predicates
encoded -- the population of out-of-coverage cases shrinks and the
\emph{error-resolution} load decreases. The CRA, operating on a shrinking
residual, imposes progressively less reactive review burden. This does
\emph{not} mean human involvement trends to zero: the foundational
accountability and context-governance function
(Section~\ref{subsec:non_delegability}) is invariant and per-transaction. What
a maturing ontology changes is the \emph{mix} of human work --- less time
spent resolving misclassifications, with no reduction in the legal
responsibility the human carries for the record. The coverage boundary itself
never reaches zero either; novel transactions, regulatory changes, and
jurisdictional edge cases continually renew the residual. The design accepts
both realities and manages them, rather than claiming to eliminate the human.

\subsection{Future Extension: Per-Client Determinism (ADR-011)}
\label{subsec:adr011_preview}

The harness principle admits a natural extension to the per-client layer.
A Client-Specific Determinism Agent (proposed in ADR-011) would observe an
individual client's transaction history, identify recurring patterns currently
resolved via Tier 3 inference, and propose client-specific Tier 1 or Tier 2
rules for those patterns. This extends the harness locally, without modifying
the shared ontology.

The same co-responsibility constraint applies: no client-specific rule is
activated without explicit human co-signature from the client's designated
accountant. The same data sovereignty constraint applies: the agent operates
entirely on client infrastructure.

ADR-011 is explicitly deferred to Phase~3+ and is documented here to
establish its architectural lineage: it is an application of the same harness
design principle at a finer (per-client) granularity, not a new concept.
The iterative loop of ADR-009 -- continuously identifying decisions that can
be migrated from inference to deterministic logic -- extends naturally and
explicitly to the client layer once production deployments provide the
empirical base required.

\subsection{Neurosymbolic Coupling Maturity: Positioning Kontablo}
\label{subsec:coupling_maturity}

Luong Tuan \& Sanyal (2026) propose a six-level maturity model for
neurosymbolic coupling in enterprise AI systems:

\begin{description}
  \item[L0 -- Ungrounded] LLM operates without ontological context.
  \item[L1 -- Context-Injected] Ontology provides prompt context (input-side).
  \item[L2 -- Discovery-Constrained] Ontology filters available tools and skills.
  \item[L3 -- Process-Gated] Ontology enforces approval gates and escalation during execution.
  \item[L4 -- Output-Validated] Ontology validates and constrains LLM outputs post-generation.
  \item[L5 -- Closed-Loop] Full bidirectional coupling: ontology constrains inputs, guides reasoning, validates outputs, and evolves from agent experience.
\end{description}

Kontablo v0.1.0 operates at \textbf{L2--L3}: the Three-Tier Resolution
Pipeline restricts tool discovery to valid UUID space (L2), and the
Co-responsibility Architecture enforces human co-signature and audit trail
for low-confidence outputs (L3). The ontology-as-constraint mechanism
described in Section~\ref{subsec:ontology_constraint} -- UUID membership
validation before ledger write -- is a hard L2 enforcement, not merely a
probabilistic guardrail.

\textbf{L4--L5 are explicitly deferred to future work.} A full output-side
validator would check that the semantic reasoning chain leading to a UUID is
ontologically consistent, not merely that the UUID itself exists. This
requires translating the agent's reasoning path into a machine-checkable form
against the Domain Ontology -- an active research problem (Luong Tuan \&
Sanyal, 2026, Section~7). The architectural lineage for this extension is
clear, and it is the natural next step once the L2--L3 production baseline is
validated by the CPA community feedback the SSRN/Zenodo release is intended
to solicit. Framing Kontablo at L2--L3 is not a concession but a precision:
it allows the specific contribution of each level to be evaluated
independently, and it places output-side validation in the correct temporal
and engineering sequence.
