\section{Related Problems and Convergent Prior Art}

The translation mechanism of the previous section is not unique to accounting. Several
mature fields solve the same problem --- reconcile two structures that share no common
coordinate system --- by the same move: lift to a richer intermediate structure, operate
there, project back. We position Kontablo against this prior art at three explicit strengths
of claim, so that a rigorous instance is never mistaken for a suggestive metaphor.

\subsection{Foundation: constructions Kontablo instantiates}
Three bodies of work supply the actual machinery and are cited as ground truth: functorial
data migration (Spivak, 2012) and ologs (Spivak \& Kent, 2012) for the schema-to-schema
translation; cellular-sheaf cohomology for data fusion (Robinson, 2017; Curry, 2014;
Hansen \& Ghrist, 2019) for the local-to-global reconciliation obstruction; and the Pacioli
group with its multi-dimensional generalization (Ellerman, 2014) for the accounting-native
algebra. These are not analogies --- Kontablo's import/export and reconciliation are
instances of them.

\subsection{Sibling: fields solving the same problem the same way}
Ontology and schema matching in the semantic web (Thiéblin et al., 2020) target precisely
the heterogeneity Kontablo calls Babel, and the Ontology Alignment Evaluation Initiative
provides standardized benchmarks worth borrowing for evaluation. The most literal twin is
\emph{phylogenetic reconciliation}: a gene tree and a species tree, distinct in topology, are
reconciled by embedding one into the other and explaining the discordance by discrete events
(duplication, loss, transfer), with a rich complexity theory of when the reconciliation is
tractable (e.g., Hasić \& Tannier, 2019). This is the tree-to-tree-via-richer-structure
problem with a different vocabulary, and its cost models are a direct precedent for
quantifying linearization loss. Interlingua machine translation (Vauquois, 1968) supplies the
pivot architecture itself.

\subsection{Machine learning: structural alignment as optimization}
A direct line of machine-learning work attacks the same alignment problem and is the most
likely source of practical leverage. Gromov--Wasserstein optimal transport (Mémoli, 2011;
Xu et al., 2019) aligns two relational structures that live in \emph{incomparable} spaces by
matching their internal distance structure alone, without any shared feature coordinates ---
exactly the situation of two charts of accounts with no common key. It yields a coupling
between the nodes of a source structure and a target structure as the solution of a
well-posed optimization problem.

Its role in Kontablo is deliberately bounded and subordinate to ADR~009. Gromov--Wasserstein
alignment is a candidate for \emph{mapping-proposal generation under sparse labels}: an
optimization-based, reproducible alternative to stochastic language-model guessing when no
deterministic rule yet exists for a given account. It is never the mapping authority. The
optimal coupling is a proposal that must pass the deterministic boundary and human
disposition; production mapping remains deterministic and auditable.

\subsection{Neurosymbolic AI and Ontology-Constrained Enterprise Agents}

A growing body of literature on neurosymbolic AI --- integration of formal
symbolic reasoning with neural language models --- directly converges on the
design principles underlying Kontablo's harness architecture. We position
Kontablo within this literature at the level of application and contribution.

\textbf{Ontological grounding reduces financial hallucination empirically.}
Luong Tuan \& Sanyal (2026) present a controlled experiment (1,800 runs, five
regulated industries, three LLMs) demonstrating that ontology-coupled agents
significantly outperform ungrounded baselines on metric accuracy and role
consistency. Independently, Agand (2026) introduces a
\emph{Universal Fact Ledger} --- deterministic typed variables grounding
financial agent reasoning --- and shows this compresses financial hallucination
to 1.2\%, with residual error concentrating at domain boundaries. Both
findings are direct empirical support for Kontablo's architectural
Principle~5 (determinism over stochasticity wherever movable).

\textbf{Ontologies provide categorically different guarantees than RAG.}
Sharma et al. (OG-RAG, EMNLP 2025) demonstrate $+55\%$ fact recall from
ontology-anchored hypergraph retrieval, yet note that formal constraint
enforcement requires machine-readable ontological definitions that
document-chunk retrieval cannot supply. We develop this categorical distinction
--- retrieval improves the odds; ontology bounds the output space --- in
Section~\ref{subsec:ontology_vs_rag}.

\textbf{Kontablo's contribution relative to this literature.}
Existing neurosymbolic enterprise architectures (Luong Tuan \& Sanyal, 2026;
Foundation AgenticOS) address general enterprise domains (fintech, healthcare,
insurance) with industry-specific ontologies. Kontablo specializes this
architecture to \emph{universal accounting} with three contributions not
present in the general case: (1) a UUID-keyed graph that is jurisdictionally
composable across all 195 sovereign states, (2) deterministic boundary
predicates that enforce accounting-native invariants (debit/credit nature,
IAS~29 restatement rules, Zakat base exclusions) independently of the
language model, and (3) an explicit universality proof via the IFRS-pure
exhaustion argument (Appendix~D) that the coverage boundary is finite and
structured, not open-ended.

\subsection{Open accounting problems the same mechanism addresses}
Framing the mechanism formally widens its surface of application. Three open problems in
global accounting reduce to the same lift--reconcile--project pattern. \textbf{Consolidation
and intercompany elimination} is a gluing of local ledgers into a global one --- the precise
scenario for which the sheaf-cohomological obstruction $H^{1}$ was introduced, so a failed
consolidation has a computable, localizable cause. \textbf{Multi-GAAP reconciliation}
(IFRS~$\leftrightarrow$~US~GAAP~$\leftrightarrow$~local) is span completion through the
universal pivot, with the residual obstruction marking which reconciliations are impossible
without loss. \textbf{Chart-of-accounts versioning and provenance} is functorial migration
across time: a schema mapping between an old and a new chart induces the migration functors
that move historical data and queries consistently (Spivak, 2012). Each is a candidate
spin-out result rather than a claim the present paper proves.
