\subsection{Remarks: the boundaries of these results}
\label{sec:remarks}

The following remarks state, in one place and without hedging, the limits of what
the evaluation above establishes. They are not disclaimers appended for caution;
they are load-bearing scope conditions, and any summary of this work that omits
them misrepresents it.

\paragraph{Remark I (Synthetic data).} The consolidation validation runs on
\emph{synthetic} trial balances constructed from the ontology's own committed code
sets. No codes are fabricated and placeholders are excluded from Tier-1. The run
demonstrates the architecture and the resolution behaviour; it is not a
measurement of real-world ledger data, and no claim of real-ledger validation is
made.

\paragraph{Remark II (56 of 60).} Tier-1 coverage is exercised against
primary-source-cited statutory charts in \emph{56 of the 60} statutory-chart
jurisdictions, not all 60. The four unexercised overlays are mapped at the
universal IFRS-anchored layer but not yet checked against a primary-source chart.

\paragraph{Remark III (Estimates, not measurements).} The core-coverage figures
($\sim$94\% by posting count, $\sim$99\% with the extended core) are model-based
estimates from a committed, labeled synthetic frequency distribution, regenerable
from \pathtt{scripts/coverage_benchmark.py}. They are not measurements of real
ledgers, and the earlier ``benchmarked against thousands of SME ledger exports''
framing has been retired.

\paragraph{Remark IV (The human is always principal).} The Co-responsibility
Architecture keeps legal accountability with the human operator at all times. High
model confidence changes the \emph{friction} of review --- how much scrutiny a
proposal invites --- never its \emph{presence}. The agent never decides alone, and
this paper does not describe it as doing so.

\paragraph{Remark V (Prototype-grade implementation).} The reference
implementation is prototype-grade relative to this paper: it demonstrates the
architecture, not the validated product. REST is fully implemented; the gRPC
surface implements only the deterministic RPCs (account queries, mapping,
consolidation with intercompany elimination, balance validation), with
LLM-dependent RPCs returning \texttt{UNIMPLEMENTED}. gRPC should be read as
``deterministic core implemented, remainder planned,'' not as feature parity with
REST.
